---
title: Generate Reproducible & Live HTML and PDF Conference Posters Using RMarkdown
author:
  - name: Anton S. Hesse
    affil: 1
    orcid: '0000-0001-8456-7343'
  - name: Christopher J. Lundstrom
    orcid: '0000-0002-1527-1685'
    affil: 1
affiliation:
  - num: 1
    address: University of Minnesota-Twin Cities, School of Kinesiology
column_numbers: 3
logoright_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
logoleft_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: packages.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Abstract

## Purpose

Cardiopulmonary exercise testing (CPET) is the gold standard for
non-invasively measuring cardiorespiratory fitness and is popular in
many fields. CPET analyzes expired gases, usually generating one data
point per breath (breath-by-breath). Consequentially, this data requires
processing before later analysis because respiratory variability is much
higher than the underlying metabolism it reflects. There is no universal
method to process breath-by-breath data, and previous research shows
that different approaches can affect downstream results. Moreover, there
is little research on the popularity of different data processing
methods and how often they are described in peer-reviewed literature.
This scoping review aims to document the popularity and prevalence of
CPET data processing methods.

## Methods

We followed current scoping review methodology and searched three
databases for papers whose methods included CPET with gas exchange. We
screened those results using regular expressions and machine learning to
identify papers where authors collected gas data breath-by-breath. Using
regular expressions, we documented which, if any, gas data processing
procedures were described in the methods section regarding outlier
removal, interpolation, and averaging methods.

## Results

Of the 8420 articles analyzed, 377 (4.5%) and 583 (6.9%) described
outlier removal and interpolation, respectively. An estimated 66.8 ±
2.8% reported averaging methods (n = 1078). When documented, the most
popular outlier cutoffs are mean ± 3 or 4 SD (39.0% and 51.7%,
respectively), and the dominating interpolation time frame and procedure
were one second (94.0%) and linear interpolation (92.2%), respectively.
When described, the most popular averaging methods are 30 (30.8%), 60
(12.4%), 15 (11.6%), 10 (11.0%), and 20 (8.1%) second bin averages.

## Conclusions

Despite the known impact on later analyses, only two-thirds of studies
using breath-by-breath data clearly documented their averaging methods.
Few papers explicitly describe gas data processing regarding outlier
removal and interpolation. The importance of data processing depends on
the subsequent analysis, but better documentation can improve
reproducibility and study comparisons.

# Introduction

-   Some combination of outlier removal, interpolation, and data
    averaging is typically required before calculating VO2max,
    ventilatory thresholds, and VO2 kinetics from CPETs
-   Prior research has shown data averaging duration affects calculated
    VO2max [@robergs2003; @robergs2010], but less is known about how
    outlier removal and interpolation affect this and other values
-   Before investigating how data processing decisions affect later
    outcomes, it is important to determine the prevalence of these
    procedures when reported
-   Quantifying the reporting percentage of data processing decisions
    improves study reproducibility

# Methods

-   This scoping review followed the Preferred Reporting Items for
    Systematic reviews and Meta-Analyses extension for Scoping Reviews

-   We downloaded a list of original, peer-reviewed articles from 3
    databases that likely included original research on human subjects
    with breath-by-breath data collection and an exercise test

-   Downloaded full-text articles were converted articles to .txt format
    and the text was normalized easier analysis

-   We used regular expressions and machine learning classification
    algorithms to remove articles not satisfying the required criteria

-   We extracted text regarding outlier, interpolation, and data
    averaging methods using regular expressions and quantified their
    reporting frequency and prevalence when reported

This package uses the same workflow approach as the R Markdown you know
and love. Basically it goes from RMarkdown \> Knitr \> Markdown \>
Pandoc \> HTML/CSS \> PDF. You can even use the bibliography the same
way [@R-posterdown].

# Results

Usually you want to have a nice table displaying some important results
that you have calculated. In `posterdown` this is as easy as using the
`kable` table formatting you are probably use to as per typical R
Markdown formatting.

You can reference tables like so: Table \@ref(tab:mytable). Lorem ipsum
dolor sit amet, consectetur adipiscing elit. Aliquam placerat augue at
velit tincidunt semper. Donec elementum porta posuere. Nullam interdum,
odio at tincidunt feugiat, turpis nisi blandit eros, eu posuere risus
felis non quam. Nam eget lorem odio. Duis et aliquet orci. Phasellus nec
viverra est.

```{r mytable, out.width='80%'}
knitr::kable(iris[1:10, 1:4], caption = 'Table caption.',align = 'c',"html")
```

Or with figures: Figure \@ref(fig:standard-plot), or Figure
\@ref(fig:morefigs).

```{r standard-plot, out.width='80%', fig.align='center', fig.cap='Great figure!', fig.height=5}
plot(mtcars[1:2])
```

```{r morefigs, out.width='80%', echo=TRUE, fig.cap='Amazing, right?!', fig.height=5}
data <- iris

plot(x = data$Sepal.Length, 
     y = data$Sepal.Width, 
     col = data$Species,
     pch = 19, 
     xlab = "Sepal Length (cm)",
     ylab = "Sepal Width (cm)")

```

# Next Steps

Aliquam sed faucibus risus, quis efficitur erat. Vestibulum semper
mauris quis tempus eleifend. Aliquam sagittis dictum ipsum, quis viverra
ligula eleifend ut. Curabitur sagittis vitae arcu eget faucibus. In non
elementum felis. Duis et aliquam nunc. Nunc pulvinar sapien nunc, vel
pretium nisi efficitur in. Fusce fringilla maximus leo et maximus. Fusce
at ligula laoreet, iaculis mi at, auctor odio. Praesent sed elementum
justo. Aenean consectetur risus rhoncus tincidunt efficitur. Praesent
dictum mauris at diam maximus maximus [@R-posterdown].

# Conclusion

Try `posterdown` out! Hopefully you like it!

```{r, include=FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```

# References
